Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-14 pecomplexity.tex
editing results more

\newtheorem*{mydef}{Definition}
\title{Determinism, Complexity, and Predictability in Computer Performance\\
Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
\\On Quantifying Predictability through Structural Complexity
\\On Quantifying Predictability with Structural Complexity
}

@@

As we stated in Section \ref{sec:intro} we intend to show two primary contributions of this work:

[[Joshua: Probably need to rephrase these]]
\begin{enumerate}

\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)

\subitem WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

@@

\subitem We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

\subitem This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.

@@

 In this section we will validate these two key findings.

@@

As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we will explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
We find that the relationship between the two is roughly linear. The single
outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
namely that in the absence of any large features, as all the other regimes have,
the WPE effectively falls back to the standard PE which the noisy behavior
drives toward 1.0.

@@

\begin{itemize}
\item Complexity need not be hard to predict (can point at the simple predictions paper)
\item random walk for example is best predicted by guess what just happend
\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not. 
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE

\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction

\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working.

Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}. [[Joshua: Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable. e.g., white noise and deterministic chaos are both complex but one is predictable. The kind of complexity is important. Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]
\begin{figure}[htbp]
  \centering
       \caption{
For each of
    these, the word length used is $6$.       
       The MASE values for LMA against ARIMA. The dashed line is the
    identity, delineating the traces for which either LMA or ARIMA performed
   better. All traces except those from {\tt SVD$_1$} lie above the line,
  indicating that LMA is better suited prediction method for the %computer
   traces considered.}\label{fig:lma_vs_arima}  
  \begin{subfigure}{\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. }
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \\
  \begin{subfigure}{\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
   % identity, delineating the traces for which either LMA or ARIMA %performed
%    better. All traces except those from {\tt SVD$_1$} lie above the line,
%    indicating that LMA is better suited prediction method for the %computer
%   traces considered.}\label{fig:lma_vs_arima}

  %\end{subfigure}
\end{figure} 
In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

@@

%%Now switch form the existence of structure as signaled by WPE to the type of structure

>>>>

\newtheorem*{mydef}{Definition}
\title{Determinism, Complexity, and Predictability in Computer Performance\\
Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
\\ Quantifying Predictability through Structural Complexity
\\ Quantifying Predictability using Structural Complexity
}

@@

BRAINSTORMING 
\begin{itemize}
\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
\item random walk for example is best predicted by guess what just happend[[move to introduction]]
\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:

\begin{enumerate}

\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.

\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

@@

\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.

@@

@@

%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]

As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is very non-trivial. As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do. Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.

In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}. We will not further validate this claim through some examples.

To further validate this finding we present an in-depth analysis of three examples (\col, \svd$_5$ and \gcc ) which nicely cover the range of complexity from low to high respectively. In Figures~\ref{fig:lma_vs_arima} and \ref{fig:arima_pred_vs_ent} we plot the MASE of LMA and ARIMA (respectively) against the weighted permutation entropy of each of the 15 runs of all 8 programs.

  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. }
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
\caption{
For each of these, the word length used is $6$. The MASE values for LMA against ARIMA. The dashed line is the identity, delineating the traces for which either LMA or ARIMA performed better. All traces except those from \svd$_1$ lie above the line, indicating that LMA is better suited prediction method for the computer traces considered.}\label{fig:lma_vs_arima}  
\end{figure}

%\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

%\item \also do \svd$_5$

%\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

We find that the relationship between the two is roughly linear. The single
outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
namely that in the absence of any large features, as all the other regimes have,
the WPE effectively falls back to the standard PE which the noisy behavior
drives toward 1.0.

@@

Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}. [[Joshua: This is not really true ... at least by these definitions because it doesn't really explain "hard to predict" ...Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable. e.g., white noise and deterministic chaos are both complex but one is predictable. The kind of complexity is important. Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]

In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as proposed in \cite{autoArima} and say we got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and too little forward information transfer is present we may not be able to do better than the random walk, in which case we should not worry ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}) we can most likely do much better than the random walk and should search for more adequate prediction strategies.

@@
